---
title: "US Mineral Commodity Summary (2022 - 2018)"
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
---

```{r setup, include=FALSE}
# Libraries Used:

# install.packages("remotes")
# remotes::install_github("hrbrmstr/albersusa", build_vignettes = TRUE)
library(flexdashboard) # For dashboard 
library(echarts4r) # For line and bar charts of Chart C and D
library(ggplot2) # For plotting map and bar chart
library(ggiraph) # For Interactive plots of Chart A (added tooltips)
library(dplyr) # For data manipulation 
library(tidyverse) # For data manipulation
library(patchwork) # For arranging plots on the plot canvas
library(albersusa) # For US map 
# library(rnaturalearth)
# library(rnaturalearthdata)
# library(rnaturalearthhires)
library(sf)  # For working with sf objects
library(ggtext) # For element_markdown() used for alignment
library(ggpubr) # For ggarrange() used for combining plots on canvas (similar to faceting)
library("cowplot") # For ggdraw()
library(mapview) # For Chart B map
#library(leaflet.extras2) # For Chart B map
```

Page 1
===================================== 

column {data-width=650}
-----------------------------------------------------------------------

### 2022 Mineral Production per State 

```{r}
# Data creation
Minerals <- data.frame (
  States = c("Arizona" ,"Nevada" ,"Texas" ,"California" ,"Minnesota" ,"Alaska" ,"Florida" ,"Utah" ,"Michigan" ,"Missouri" ,"Wyoming" ,"Georgia" ,"Pennsylvania" ,"Ohio" ,"New York"  ,"Tennessee" ,"Alabama" ,"North Carolina" ,"Colorado" ,"Wisconsin" ,"Montana" ,"Virginia" ,"South Carolina" ,"New Mexico" ,"Illinois" ,"Indiana" ,"Kansas" ,"Kentucky" ,"Arkansas" ,"Oklahoma" ,"Louisiana" ,"Washington" ,"Idaho" ,"Iowa" ,"Maryland" ,"Oregon" ,"South Dakota" ,"New Jersey" ,"Nebraska" ,"West Virginia" ,"Massachusetts" ,"Mississippi" ,"Connecticut" ,"Hawaii" ,"New Hampshire" ,"Vermont" ,"Maine","North Dakota" ,"Rhode Island" ,"Delaware"),
  Values = c(10100 ,8930 ,8030 ,5610 ,4780 ,4510 ,2810 ,3600 ,3360 ,3150 ,2480 ,2320 ,2060 ,1490 ,1950 ,1940 ,1920 ,1900 ,1870 ,1720 ,1600 ,1530 ,1160 ,1470 ,1250 ,1380 ,1170 ,806 ,1100 ,1030 ,1030 ,901 ,371 ,846 ,414 ,693 ,475 ,425 ,256 ,204 ,206 ,225 ,194 ,156 ,135 ,136 ,100 ,105 ,88 ,25),
  Rank = c(1 ,2 ,3 ,4 ,5 ,6 ,7 ,8 ,9 ,10 ,11 ,12 ,13 ,14 ,15 ,16 ,17 ,18 ,19 ,20 ,21 ,22 ,23 ,24 ,25 ,26 ,27 ,28 ,29 ,30 ,31 ,32 ,33 ,34 ,35 ,36 ,37 ,38 ,39 ,40 ,41 ,42 ,43 ,44 ,45 ,46 ,47 ,48 ,49 ,50)
)

# Remove trailing spaces in States column
Minerals$States <- str_trim(Minerals$States, "right")

# Create a percentage column showing each State's contribution to the nation's total
Minerals <- Minerals %>% mutate(Percentage = paste0(round(Values / sum(Values) * 100, 1), "%"))

# Create new column for tooltip
Minerals <- Minerals %>% mutate(tooltip_text = paste0(toupper(States), "\n", "$ ", formatC(Values, format = "d", big.mark = ","), "\n", Percentage))

# Create the bar chart 
bar_plot <- ggplot(Minerals, aes(x = reorder(States, Values), y = Values,
                     tooltip = tooltip_text, data_id = States)) + 
  geom_col_interactive(color = "grey", fill = "#0072b2", size = 0.5) + 
  theme_minimal() +
  theme(axis.text = element_text(size = 12), 
        legend.position = "none",
        plot.caption = element_markdown(lineheight = 1.2),
        plot.title = element_textbox_simple(
          hjust = 0,
          vjust = 0,
          size = 15,
          padding = margin(5.5, 5.5, 5.5, 5.5),
          margin = margin(0, 0, 5.5, 0)
        )) + 
  scale_y_continuous(label = scales::dollar_format()) +
  labs(title = "<span style = 'font-size:14pt'> <b> US States Ranked By Mineral Production Value (2022) </b> </span>",
       caption = "<span style = 'font-size:11pt'> <br>**Arizona** made up more than 10% of U.S. mineral production value, and is the leading<br> copper-producing state making up around 70% of domestic copper production <br> <br> 
       **Nevada** remains the nation's top producer of gold and silver,<br> along with being the only lithium-producing state</span>") +
  ylab("USD in Mil. $") + 
  xlab("") +
  coord_flip()

# Get US state data 
us_sf <- usa_sf("lcc") %>% 
  mutate(State = as.character(name))

# Join Our Minerals df with US map df 
Map_data <- left_join(us_sf, Minerals, by = c("State" = "States"), keep = TRUE)

# Create map plot
map_plot <- ggplot() + 
  geom_sf_interactive(data = Map_data, size = 0.125, aes(fill = NULL, data_id = State, tooltip = tooltip_text), color = "white") + 
  theme_void() + 
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5, vjust = 1)) +
  labs(title = "<span style = 'font-size:11pt'>   The **Western region** of the US<br>   accounted for more than<br>   **one-third** of US minerals<br>   production value at $35.9 billion </span>",
    caption = "<span style = 'font-size:12pt'> The **Southern Region** was the top<br> producer of industrial minerals<br> with a value of $13.2 billion </span>") +
  theme(
    plot.caption = element_markdown(lineheight = 1.2),
    plot.title.position = "plot", 
    plot.title = element_textbox_simple(lineheight = 0.8, size = 13,
                                        padding = margin(5.5, 5.5, 5.5, 5.5),
                                        margin = margin(0, 0, 5.5, 0)))
  
# Draw bar plot and map plot on same canvas 
figure2 <- ggdraw() +
  draw_plot(map_plot, x = 0, y = .25, width = .45, height = .6)+
  draw_plot(bar_plot, x = .45, y = 0.01, width = .55, height = 0.99)

# Interactive bar chart and US map
p <- girafe(ggobj = figure2, 
            options = list(
              opts_hover(css = ''),
              opts_hover_inv(css = "opacity:0.5;"), 
              opts_sizing(rescale = TRUE)
            ),
            height_svg = 10,
            width_svg = 16) %>% 
  girafe_options(opts_hover(css = "fill:cyan;"))

p

```

> The U.S. produced $98.2 billion worth of nonfuel minerals in 2022. Nonfuel mineral production is categorized into two main categories by the USGS, metals/metallic minerals and industrial minerals. While $34.7 billion worth of metals were produced in 2022, industrial mineral production value was nearly double at $63.5 billion.

Column {data-width=350}
-----------------------------------------------------------------------

### Past 5 years U.S. Mineral Industry Trends {.no-title}

```{r}
# Get USGS2 (2023) into R
USMineralTrends <- read.csv("C:/Users/admin/Desktop/Data Viz Final project/widetolong.csv") 

## Data Reference : USGS (U.S. Geological Survey) (2023). *MINERAL COMMODITY SUMMARIES 2023*, USGS Publications Repository Website, accessed 4th June 2023. https://pubs.usgs.gov/periodicals/mcs2023/mcs2023.pdf, page 9, *Table 1.—U.S. Mineral Industry Trends*

## First Trend chart - Mine Production over past 5 years

# Filter Mine Production Category, transform long to wide and convert a column as character
USMineralTrends_MineCategory <- USMineralTrends %>% filter( MainFilter == "Total mine production (million dollars)")
USMineralTrends_MineCategory_spread <- spread(USMineralTrends_MineCategory, Subfilter, Value) %>% mutate(across(c(Coal, `Industrial minerals`, Metals), ~ . / 1000))
USMineralTrends_MineCategory_spread$Year <- as.character(USMineralTrends_MineCategory_spread$Year)


# Create Trend chart using echarts4r package
e1 <- USMineralTrends_MineCategory_spread %>% 
  e_charts(x = Year, height = '40vh', width = 'auto', elementId = "chart1") %>%  # Use relative height
  e_line(serie = Coal) %>% 
  e_line(serie = `Industrial minerals`) %>% 
  e_line(serie = Metals) %>% 
  e_tooltip(trigger = "axis",formatter = e_tooltip_pointer_formatter("currency")) %>% 
  e_format_y_axis(suffix = " Bil. $") %>% 
  e_grid(left = '15%', right = '5%') %>% 
  e_hide_grid_lines(which = c("x")) %>% 
  e_legend(orient = 'vertical', right  = '0', top = '10%', show = FALSE) %>%
  e_title("U.S. Mineral Industry Trends", subtext = "Mine Production over past 5 years", left = "center" ) 

## Second Trend chart - Employment over past 5 years

# Same process
USMineralTrends_Employment <- USMineralTrends %>% filter( MainFilter == "Employment (thousands of workers)")
USMineralTrends_Employment_spread <- spread(USMineralTrends_Employment, Subfilter, Value)
USMineralTrends_Employment_spread$Year <- as.character(USMineralTrends_Employment_spread$Year)

e2 <- USMineralTrends_Employment_spread %>% 
  e_charts(x = Year, height = '40vh', width = 'auto',elementId = "chart2") %>%  # Use relative height
  e_line(serie = `Coal mining`) %>% 
  e_line(serie = `Nonfuel mineral mining`) %>% 
  e_line(serie = `Chemicals and allied products`) %>% 
  e_line(serie = `Stone, clay, and glass products`) %>% 
  e_line(serie = `Primary metal industries`) %>% 
  e_tooltip(trigger = "axis") %>% 
  e_format_y_axis(suffix = "k") %>% 
  e_grid(left = '15%', right = '5%') %>% 
  e_hide_grid_lines(which = c("x")) %>% 
  e_legend(orient = 'vertical', right  = '0', top = '15%', show = FALSE) %>%
  e_title(subtext = "Employment over past 5 years (in )", left = "center", top = 10 ) 

## Third Trend chart - Weekly Earnings over past 5 years

# Same process
USMineralTrends_Earnings <- USMineralTrends %>% filter( MainFilter == "Average weekly earnings of workers (dollars)")
USMineralTrends_Earnings_spread <- spread(USMineralTrends_Earnings, Subfilter, Value)
USMineralTrends_Earnings_spread$Year <- as.character(USMineralTrends_Earnings_spread$Year)

e3 <- USMineralTrends_Earnings_spread %>% 
  e_charts(x = Year, height = '40vh', width = 'auto',elementId = "chart3") %>%  # Use relative height
  e_line(serie = `Coal mining`) %>% 
  e_line(serie = `Chemicals and allied products`) %>% 
  e_line(serie = `Stone, clay, and glass products`) %>% 
  e_line(serie = `Primary metal industries`) %>% 
  e_tooltip(trigger = "axis") %>% 
  e_format_y_axis(suffix = " $") %>% 
  e_grid(left = '15%', right = '5%') %>% 
  e_hide_grid_lines(which = c("x")) %>% 
  e_legend(orient = 'vertical', right  = '0', top = '15%', show = FALSE) %>%
  e_title(subtext = "Weekly Earnings over past 5 years", left = "center", top = 10 ) %>% 
  e_connect(c("chart1", "chart2")) 

# Combine all 3 trend charts
e_arrange(e1, e2, e3)
```

> The total value of industrial minerals production was **$63.5 billion**, a 10% increase from that in 2021. Of this total, $31.4 billion was construction aggregates production (construction sand and gravel and crushed stone). **Crushed stone** was the leading nonfuel mineral commodity in 2022 with a production value of $21 billion and accounted for 21% of the total value of U.S. nonfuel mineral production.

Page 2
===================================== 

column {data-width=650}
-----------------------------------------------------------------------

### Compare Mineral Production per State - 2022 vs 2018 

```{r}
# Create USGS1 (2023) and USGS (2019) in R
Minerals <- data.frame (
  States = c("Arizona" ,"Nevada" ,"Texas" ,"California" ,"Minnesota" ,"Alaska" ,"Florida" ,"Utah" ,"Michigan" ,"Missouri" ,"Wyoming" ,"Georgia" ,"Pennsylvania" ,"Ohio" ,"New York"  ,"Tennessee" ,"Alabama" ,"North Carolina" ,"Colorado" ,"Wisconsin" ,"Montana" ,"Virginia" ,"South Carolina" ,"New Mexico" ,"Illinois" ,"Indiana" ,"Kansas" ,"Kentucky" ,"Arkansas" ,"Oklahoma" ,"Louisiana" ,"Washington" ,"Idaho" ,"Iowa" ,"Maryland" ,"Oregon" ,"South Dakota" ,"New Jersey" ,"Nebraska" ,"West Virginia" ,"Massachusetts" ,"Mississippi" ,"Connecticut" ,"Hawaii" ,"New Hampshire" ,"Vermont" ,"Maine","North Dakota" ,"Rhode Island" ,"Delaware"),
  Values = c(10100 ,8930 ,8030 ,5610 ,4780 ,4510 ,2810 ,3600 ,3360 ,3150 ,2480 ,2320 ,2060 ,1490 ,1950 ,1940 ,1920 ,1900 ,1870 ,1720 ,1600 ,1530 ,1160 ,1470 ,1250 ,1380 ,1170 ,806    ,1100 ,1030 ,1030 ,901 ,371 ,846 ,414 ,693 ,475 ,425 ,256 ,204 ,206 ,225 ,194 ,156 ,135 ,136 ,100 ,105 ,88 ,25),
  Values2018 = c(1450, 3440, 6690, 903, 4560, 1380, 200, 31, 3550, 1960, 141, 208, 1780, 1060, 680, 672, 513, 536, 135, 410, 340, 2470, 4050, 404, 2930, 1130, 209, 7880, 162, 295, 1160, 1790, 1210, 136, 1200, 894, 530, 1920, 56, 1050, 339, 1460, 6030, 2940, 104, 1290, 1090, 263, 2730, 2410),
  Rank = c(1 ,2 ,3 ,4 ,5 ,6 ,7 ,8 ,9 ,10 ,11 ,12 ,13 ,14 ,15 ,16 ,17 ,18 ,19 ,20 ,21 ,22 ,23 ,24 ,25 ,26 ,27 ,28 ,29 ,30 ,31 ,32 ,33 ,34 ,35 ,36 ,37 ,38 ,39 ,40 ,41 ,42 ,43 ,44 ,45  ,46 ,47 ,48 ,49 ,50),
  Rank2018 = c(20, 7, 2, 1, 4, 21, 43, 50, 6, 13, 47, 35, 15, 27, 29, 28, 32, 34, 44, 33, 40, 11, 5, 38, 9, 24, 39, 1, 45, 42, 23, 16, 18, 46, 17, 31, 36, 14, 49, 26, 41, 19, 3, 8, 48, 22, 25, 37, 10, 12))

## Data Reference : USGS (U.S. Geological Survey) (2019). *MINERAL COMMODITY SUMMARIES 2019*, USGS Publications Repository Website, accessed 4th June 2023. https://d9-wret.s3.us-west-2.amazonaws.com/assets/palladium/production/atoms/files/mcs2019_all.pdf, pages 10-11, *TABLE 3.—VALUE OF NONFUEL MINERAL PRODUCTION IN THE UNITED STATES AND PRINCIPAL NONFUEL MINERALS PRODUCED IN 2018*

# Remove trailing spaces in States column
Minerals$States <- str_trim(Minerals$States, "right") #library(tidyverse) # For data manipulation


# Create a percentage columns showing each State's contribution to the nation's total, Change in mineral production value from 2018 to 2022, and other columns for tooltips.
Minerals <- Minerals %>% mutate(Percentage_22 = paste0(round(Values/sum(Values)*100,1),"%"),
                                Percentage_18 = paste0(round(Values2018/sum(Values2018)*100,1),"%"),
                                ValueChange = paste0("$ ",formatC(Values-Values2018, format="d", big.mark=",")),
                                TooltipState = paste0(toupper(Minerals$States)),
                                TooltipValues2022 = paste0("$ ",formatC(Minerals$Values, format="d", big.mark=",")),
                                TooltipValues2018 = paste0("$ ",formatC(Minerals$Values2018, format="d", big.mark=",")))

# Get US state data from library(albersusa)
us_sf <- usa_sf("lcc") %>% 
  mutate(State = as.character(name))

# Join Our Minerals df with US map df 
Map_data <- left_join(us_sf, Minerals, by = c("State" = "States" ), keep = TRUE)

# Create Labels
mylabel <- glue::glue("{Map_data$TooltipState} (2018)<br />{Map_data$TooltipValues2018} <br />{Map_data$Percentage_18} <br />")%>%
  lapply(htmltools::HTML)
mylabel22 <- glue::glue("{Map_data$TooltipState} (2022)<br />{Map_data$TooltipValues2022} <br />{Map_data$Percentage_22} <br />")%>%
  lapply(htmltools::HTML)

# Create tooltip (combine create tooltip columns)
mypopup <- glue::glue("<strong>{Map_data$TooltipState}</strong><br />
                      Value Increased from 2022 to 2018: {Map_data$ValueChange}") %>% 
  lapply(htmltools::HTML)

# Create map for 2018 (from 2019 summary report) data 
map2018 <- mapview(Map_data, zcol = "Percentage_18", 
                   col.regions = RColorBrewer::brewer.pal(9, "Greens"), alpha.regions = 1,
                   layer.name = "Pct 2018", 
                   popup = mypopup, label = mylabel,legend = FALSE
)

# Create map for 2022 (from 2022 summary report) data 
map2022 <- mapview(Map_data, zcol = "Percentage_22", 
                   col.regions = RColorBrewer::brewer.pal(9, "Greens"), alpha.regions = 1,
                   layer.name = "Pct 2022",
                   popup = mypopup, label = mylabel22
)

# Setting to avoid automatic colouring of maps by R
mapviewOptions("basemaps.color.shuffle" = FALSE)

# Create slider between 2022 and 2018 maps and display resulting chart/map
map2018 | map2022
```

> **CLICK [Pct 2022] TO CENTER (Button in the bottom right corner)**  </br>
> Click on States to open tooltip. Hover for data labels.

Column {data-width=350}
-----------------------------------------------------------------------

### Commodity Import Summary - 2022

```{r}

# Create USGS3 (2023) in R
SourceConsumption <- read.csv("C:/Users/admin/Desktop/Data Viz Final project/Commodity.csv") 

## Data Reference : USGS3 (U.S. Geological Survey) (2023). *MINERAL COMMODITY SUMMARIES 2023*, USGS Publications Repository Website, accessed 4th June 2023. https://pubs.usgs.gov/periodicals/mcs2023/mcs2023.pdf, page 7, *Figure 2.—2022 U.S. Net Import Reliance*

# Create bar chart using echarts4r package
SourceConsumption %>% arrange(Consumption) %>%
  #slice_tail(n = 30) %>%
  e_charts(x = Commodity, reorder = FALSE, height = '60vh') %>%
  e_bar(Consumption, bind = Source) %>% 
  e_labels(position = "right") %>% 
e_tooltip(
    formatter = htmlwidgets::JS("
      function(params){
        // Split the comma-separated string into an array
        var sources = params.name.split(',');
        var sourcesFormatted = sources.join('<br />');
        return(
          '<div style=\"font-size: 12px;\">' +
          '<strong>Imported from:</strong> <br />' + sourcesFormatted + '<br />' +
          '<strong>Commodity:</strong> ' + params.value[1] + '<br />' +
          '<strong>Consumption:</strong> ' + params.value[0] + '%' + 
          '</div>'
        );
      }
    "),
    position = htmlwidgets::JS("
      function (point, params, dom, rect, size) {
        var x = point[0];
        var y = point[1] + 10; // Shift 10px down to position the tooltip below the cursor
        return [x, y];
      }
    ")
  ) %>% 
  e_title("2022 U.S. Net Import Reliance", left = "center", top = 10) %>% 
  e_flip_coords() %>% 
  e_x_axis( 
    name = "Consumption Reliance on Imports (%)",
    nameLocation = "center",
    nameGap = 30,
    nameTextStyle = list(
      color = "grey",
      fontSize = 12
    )
  ) %>% 
  e_format_x_axis(suffix = "%") %>%
  e_toolbox_feature(
    feature = "saveAsImage",
    title = "Save as image"
  ) %>%  
  e_legend(show = FALSE) %>% 
  e_grid(left = "50%", right = "8%")%>%  # Adjust grid to make space for labels
  e_datazoom(
    type = "slider",               # Use a slider for data zoom
    start = 0,                     # Start position (0%)
    end = (20/nrow(SourceConsumption)) * 100,  # End position (20 records shown initially)
    y_index = 0                    # Apply zoom to the x-axis
  )

```

> In 2022, imports made up more than one-half of the U.S. apparent consumption for 51 nonfuel mineral commodities, and the United States was 100% net import reliant for 15 of those.